# Analysis of Fatalities, injuries and damage caused by severe weather events
## Synopsis
Based on recorded storm data from 1950 through 2011, this document tries to provide some insight into the effects of severe weather on the public. These kinds of answers can be beneficial to plan responses to severe weather events and to prepare contingency plans.

We found that convection events (Lightning, Tornadoes, Thunderstorm Wind, Hail) are the most harmful to public health. We also found that Flood events (Flash Floods, River Floods) are the most damaging to property and crops.

# Data Processing
Load the libraries we will need:
```{r}
library(ggplot2)
library(maps)
library(mapproj)
```
Our data is derived from the [NOAA Storm Database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).  
Read in the data, mapping as many fields to numerical fields as possible. We are not converting the dates at this point, as we do not need the dates in our analysis. More information about the data file is available from the National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).  
```{r}
data <- read.csv("repdata_data_StormData.csv.bz2", header = TRUE, sep = ",", 
    colClasses = c("numeric", "character", "character", "character", "numeric", 
        "character", "character", "character", "numeric", "character", "character", 
        "character", "character", "numeric", "character", "character", "character", 
        "character", "numeric", "numeric", "character", "numeric", "numeric", 
        "numeric", "numeric", "character", "numeric", "character", "character", 
        "character", "character", "numeric", "numeric", "numeric", "numeric", 
        "character", "numeric"))
```
The initial set has 902,297 observations. We first throw away all data that does not contain information we are interested in by filtering out data that did not cause fatalities, injuries or damage.  
```{r}
smallData <- data[data$FATALITIES > 0 | data$INJURIES > 0 | data$PROPDMG > 0 | 
    data$CROPDMG > 0, ]
```
This leaves us with 254,633 observations.  
The EVTYPE fields contains a large number of errors and issues. In order to report on the data, we will add an additional column named category that contains the event category as used by the NCDC: 
- convection  
- extreme temperature
- flood
- marine
- tropical cyclon
- winter
- other
This is also the order of importance with which we will treat the various events. Convection events are the most important, so this order will also decide the tie-breaker if an event belongs to more than one category. 
```{r}
convection <- c("?.*LIGHTING|LIGHTN|LIGNTNING|TORNADO|TORNDAO|WIND|TSTM|THUNDER|WND|HAIL|TURBULENCE|GUSTNADO?.*")
temperature <- c("?.*COLD|HEAT|WARM|TEMPERATURE|THERMIA?.*")
flood <- c("?.*FLOOD|RISING|STREAM FLD?.*")
marine <- c("?.*COASTAL|TSUNAMI|CURRENT|SWELLS|TIDE|WAVE|SEAS|SURF|HIGH WATER?.*")
cyclones <- c("?.*CYCLONE|HURRICANE|TYPHOON?.*")
winter <- c("?.*WINT|ICE|AVALAN|SLEET|SNOW|FREEZ|BLIZZ|FROST|GLAZE|MIXED?.*")

smallData[grepl(winter, smallData$EVTYPE, ignore.case = TRUE), "category"] = "Winter"
smallData[grepl(cyclones, smallData$EVTYPE, ignore.case = TRUE), "category"] = "Tropical Cyclones"
smallData[grepl(marine, smallData$EVTYPE, ignore.case = TRUE), "category"] = "Marine"
smallData[grepl(flood, smallData$EVTYPE, ignore.case = TRUE), "category"] = "Flood"
smallData[grepl(temperature, smallData$EVTYPE, ignore.case = TRUE), "category"] = "Extreme Temperatures"
smallData[grepl(convection, smallData$EVTYPE, ignore.case = TRUE), "category"] = "Convection"
smallData[is.na(smallData$category), "category"] = "Other"
```
The PROPDMG and CROPDMG fields need some conversion before we can do math on them. Here is the code for that. We add two extra columns that contain the property and crop damage.  

```{r}
smallData$propertydamageEXP = 1
smallData[grepl("h|H", smallData$PROPDMGEXP, ignore.case = TRUE), "propertydamageEXP"] = 100
smallData[grepl("k|K", smallData$PROPDMGEXP, ignore.case = TRUE), "propertydamageEXP"] = 1000
smallData[grepl("m|M", smallData$PROPDMGEXP, ignore.case = TRUE), "propertydamageEXP"] = 1e+06
smallData[grepl("b|B", smallData$PROPDMGEXP, ignore.case = TRUE), "propertydamageEXP"] = 1e+09
smallData$propertydamage = smallData$PROPDMG * smallData$propertydamageEXP

smallData$cropdamageEXP = 1
smallData[grepl("k|K", smallData$CROPDMGEXP, ignore.case = TRUE), "cropdamageEXP"] = 1000
smallData[grepl("m|M", smallData$CROPDMGEXP, ignore.case = TRUE), "cropdamageEXP"] = 1e+06
smallData[grepl("b|B", smallData$CROPDMGEXP, ignore.case = TRUE), "cropdamageEXP"] = 1e+09
smallData$cropdamage = smallData$CROPDMG * smallData$cropdamageEXP
```

Having gone through the steps above, we now have clean dataset containing all the information to perform exploratory data analysis.

## Exploratory Data Analysis
We begin by exploring all the variables in the data set.
### Univariate Plots

```{r}
county<-table(smallData$COUNTYNAME)
out=data.frame(item=names(unlist(county)),count=unlist(county)[],
                stringsAsFactors=FALSE)
```

```{r}
ggplot(out, aes(x=count)) + geom_histogram(binwidth=200)
```

Figure above shows the frequency graph of number of events occuring in the respective counties. The mean value of frequence of occurence is around 22, thus showing that on an average you can expect a place to have an event occuring 22 time over the years between 1950-2011. Let's have a closer look at the regions with lot of events occuring in this time span.

```{r}
maxregions<-subset(out,out$count>1500)
ggplot(maxregions, aes(x=item, y = count, fill = count))+theme(axis.text.x = element_text(angle = 90, hjust = 1))+geom_bar(stat="identity")+xlab("County Names")+ylab("Number of weather events occurred")+ggtitle("Largest number of events that have occurred in Countys")
```

The County of Washington has the highest number of weather events that has occurred over these years. Let's now have a close look as to what events have taken place most in washington as per the category we have assigned.

```{r message=FALSE, warning=FALSE}
washington2<-subset(smallData, COUNTYNAME == "WASHINGTON" & FATALITIES > 0)
washington1<-table(washington2$category)

washington<-data.frame(events=names(unlist(washington1)), count = unlist(washington1)[], stringsAsFactors = FALSE)

ggplot(washington, aes(x=events, y = count, fill = count))+geom_bar(stat="identity")+xlab("Events")+ylab("Number of occurrences")+ggtitle("Most prominent Disaster in Washington")

```

Therefore, we can conclude that the most prominent disaster in Washington is Convection. Let us now see if these were the cause of the most number of fatalities in the case of washington.

```{r}
table(washington$category)
```

Breaking our category variable into individual events, we get the following result

```{r}
table(washington2$EVTYPE)
```

## Bivariate Plots
Let's look at a few scatter plots to identify - if there is any kind of relationship present between a few features. 

We begin by comparing two features i.e fatalities and injuries. We get the following result:

```{r message=FALSE,warning=FALSE}
ggplot(smallData, aes(x=INJURIES,y=FATALITIES))+geom_point(shape=1)+geom_smooth()+scale_y_continuous(limits = c(0,100)) + scale_x_continuous(limits = c(0,1000))
```

The graph shows some kind of linear relationship between the two variables which would make sense: Let's say a person is injured due to an unforseeable calamity like flood or lightning. If this event were to have a very serious effect on the person, he/she could succumb to the injury which could lead to a fatality.

Let's look at the relationship between property damage and crop damage:

```{r message=FALSE,warning=FALSE}
fatalities<-subset(smallData,FATALITIES > 0 & INJURIES > 0)
x<-ggplot(fatalities, aes(x=PROPDMG,y=CROPDMG,))+geom_point(shape=1)+geom_smooth()
x + facet_grid(. ~ category)+scale_y_continuous(limits=c(0,1000))
```

The x-axis depicts property damage, while the y axis shows corresponding crop damage. I have faceted accroding to categories of events taking place so as to get more perspective into relationship between events.

For the large part we find that Convection events cause a lot of property damage and crop damage. this is due to the the fact that such events include cyclone and tornadoes which have a disastrous effect on both. Hence we see a smooth linear regression curve in the first grid.

Events like Extreme Temperatures have a large effect on crops rather than property as shown in the second rid. Thus, we get a perspective as to what we would be expecting in the case of economical and health damage that these weather events cause from the plot above.

##Multivariate Plots
This section will cover more detailed plots and multiple scatter plots togther to find some kind of relationships between variables quickly.

```{r message=FALSE,warning=FALSE}
ggplot(smallData, aes(LONGITUDE, LATITUDE, colour = smallData$FATALITIES,100)) + geom_point(size = 4,shape = 1, colour = "steelblue", alpha = 0.4) + labs(x = "Longitude", y = "Latitude") + scale_colour_grey() + theme_bw() 
```

Figure above shows the fatalities and in which Latitude and Longitude they are concentrated. Most of the data shows a slight amount of fatalities due to the overplotting of blue dots. In order to get a better perspective we will use the maps package to plot the fatalities on Map of USA.

```{r message=FALSE,warning=FALSE}
ggplot(smallData, aes(LONGITUDE, LATITUDE, colour = factor(cut(smallData$FATALITIES,100, label = FALSE)))) + geom_point(size = 4) + geom_point(size = 4,shape = 1, colour = "steelblue", alpha = 0.4) + labs(x = "Longitude", y = "Latitude") + scale_colour_grey() + theme_bw()+facet_grid(~category)
```

Figure above shows multiple deaths in the case of Convection Events and Flood Events.

```{r message=FALSE,warning=FALSE}
pairs(~FATALITIES+INJURIES+PROPDMG+CROPDMG,data=smallData, 
    main="Simple Scatterplot Matrix for various variables")
```

Figure above shows the relationship between various damages and fatalities. This is to find some kind of relationship between the economic and health related damages the weather events cause damage to.
Insights:
- Fatalities and crop damage have no relation whatsoever as expected.
- Injuries and Crop damage have once agin no relationship whatsover.
- Injuries and Fatalaties seems to have an almost linear relationship with property damage.
- property and crop damage see, to follow a direct relationship at some places.
On Reviewing all the insights from the graphs and exploratory data analysis, following statements can be made.
- Washington suffers from a large number of conviction events.
- Convection events cause the most damage in terms of health.
- Flooding events have the most effect on property and crops.

## Final Plots and Results
### Severe weather events that cause the largest number of incidents
Calculate a total of all fatalities and injuries, so that we can find what events have the highest number of incidents. The new column is called incidents.  

```{r}
smallData$incidents = smallData$FATALITIES + smallData$INJURIES
```

We create a new set with the aggregate of the incidents grouped by the event types.

```{r}
incidentData <- aggregate(list(incidents = smallData$incidents), by = list(event = smallData$category), 
    FUN = sum, na.rm = TRUE)
```

Here is the overview of the event categories with the number of incidents  

```{r}
incidentData$event <- reorder(incidentData$event, -incidentData$incidents)
ggplot(incidentData, aes(y = incidents)) + theme(axis.text.x = element_text(angle = 90, 
    hjust = 1)) + geom_bar(aes(x = event), data = incidentData, stat = "identity") + 
    ggtitle("Fatalities and Injuries") + xlab("Event Category") + ylab("No. of Fatalities and Injuries")
```

Clearly the convection events (Lightning, Tornadoes, Thunderstorm Wind, Hail) has the greatest effect on injury and fatality.
## Severe weather events that cause the largest amount of damage
Add a field with a total of all damage, both to property as well as to crop, so that we can find our what events cause the highest amount of damage. The new column is called damage. The column is in billions of dollars.  

```{r}
smallData$damage = ((smallData$propertydamage + smallData$cropdamage)/1e+09)
```

We create a new set with the aggregate of the damage grouped by the event types.  

```{r}
damageData <- aggregate(list(damage = smallData$damage), by = list(event = smallData$category), 
    FUN = sum, na.rm = TRUE)
```

Here is the overview of the event categories with the amount of damage  

```{r}
damageData$event <- reorder(damageData$event, -damageData$damage)
ggplot(damageData, aes(y = damage)) + theme(axis.text.x = element_text(angle = 90, 
    hjust = 1)) + geom_bar(aes(x = event), data = damageData, stat = "identity") + 
    ggtitle("Property and crop damage") + xlab("Event Category") + ylab("Amount of damage (billions of $)")
```

Clearly the flooding events (Flash Flood, River Flood) have the greatest effect on property damage and crop damage.

After Converting the data into an aggregate form and replacing state abbreviations with their name, I built a new csv containing the aggregate data and removed the states that were not present in the maps package

```{r}
states_map <- map_data("state")
data<-read.csv('events.agg.csv')
aggdata<-aggregate(data[,c('FATALITIES','INJURIES','PROPDMG','CROPDMG')],by=list(data$STATE),FUN=sum)
aggdata$Affected = aggdata$FATALITIES+aggdata$INJURIES
colnames(aggdata)<-c("STATE","FATALITIES","INJURIES","PROPDMG","CROPDMG","Affected")
aggdata$STATE<-as.character(aggdata$STATE)
p <- ggplot(aggdata, aes(map_id = STATE))
p <- p + geom_map(aes(fill = Affected), map = states_map, colour='black') + expand_limits(x = states_map$long, y = states_map$lat)
p <- p + coord_map() + theme_bw()
p <- p + labs(x = "Long", y = "Lat", title = "Health and Population damages")
print(p)
```

Figure above shows thet total damage to health i.e lives and as we can see the most affected states are the western states due to the fact that they are victims of events like tornadoes and hurricanes. Let us know look at the damage to Economy which includes property and crop damages

```{r}
aggdata$Damages<-aggdata$CROPDMG+aggdata$PROPDMG
p <- ggplot(aggdata, aes(map_id = STATE))
p <- p + geom_map(aes(fill = Damages), map = states_map, colour='black') + expand_limits(x = states_map$long, y = states_map$lat)
p <- p + coord_map() + theme_bw()
p <- p + labs(x = "Long", y = "Lat", title = "Economic Damages")
print(p)
```
As we can see the western and central states of USA have been hit hard causing huge losses to economy fro the years 1950-2011.

##Reflections
- There were many challenges in this project particularly due to the fact that there was so much cleaning up to do in order to get the data ready to be analysed. I believe, I may have ballparked a few data, however My initial EDA conforms to the final conclusion which I have given.
- A particular point about this dataset is that it contains many events which do not contribute to much damages which lead to a large redundancy of data.
- As far as the imporvement in analysis is concerned, I believe a few more correlations between the features at stake would go onto provide more insights into the data set.
- My Analysis has been that in terms of a broad sense. A few more intricate details can be covered within this data set leading to better results and conclusion from the data.
- Features are less due to the fact that a large number of values are noth fathomable and have averaged ot value.
